Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation 

### 2025-05-22 manuscript/manuscript.qmd
updated methods section

## 2. Materials and Methods

@@

### 2.1 Data Description and Preparation Overview

@@

### 2.2 Bayesian Modeling Framework

@@

*Model 2* introduces partial pooling across four groups defined by the dominant MBR numerator band; Rrs411, Rrs443, Rrs489, Rrs510. Each group receives its own slope and intercept with shared hyperpriors. On the one hand each group has its own submodel that can capture details specific to it, while the shared hyperpriors allow for the sharing of information between groups when fitting the model, thus maximizing the use of information contained in the available data. $X_i$ is the same as for Model 1, i.e. $log(MBR)$, assigned a given observation to group $i$ (@fig-eda-mbr). This model retains a constant dispersion parameter $σ$, similar to *Model 1*.
Below is the system of equations corresponding to this model, where $j$ is the group index and $i$ is the $i^{th}$ obasrvation. Mathematical description of the model can be found in the Supplementary Material. @fig-model2-struct illustrates the model structure.

@@

Model 2 - Multi-level hierarchical level with partial pooling across $MBR$ numerator groups (Rrs411, Rrs443, Rrs489, Rrs510). Model parameter prior mean and variance have Gaussian and Exponential hyperpriors. This structure allows the sharing of information across groups at model fitting time.
:::

@@

The data plotted in @fig-eda-mbr suggests heteroscedasticity - variance that is not constant. Morevoer variance seems to change with the data. Models 3 and 4 (see Supplemental Material section) investigated the role of groupwise constant heteroscedasticity and global linear heteroscedasticity, respectively. To investigate this further, Model 5 include groupwise linear heteroscedasticity. Thus,  for a given MBR group $j$ $σ_j$ is modeled as a log-linear relationship such that $$log(σ) = α_j + β_j × log(MBR)$ and then transformed via $exp(log(σ_j))$ before being passed to the likelihood. Similarly to the model mean $μ_j$, $σ_j$'s slope and intercept terms are partially pooled and so have common hyperpriors. Mathematical descriiption of the model can be found in the Supplementary Material section; @fig-model5-struct.  illustrates the model structure.

@@

::: {#fig-model6-struct}

@@

### Inference Settings

I used weakly informative priors across all models to balance flexibility and regularization:

- **Regression terms**: Normal(0, 1) unless otherwise stated  
- **Hyperpriors**: Normal and Exponential for hierarchical parameters  
- **Noise terms** (`σ`): Gamma(γ, φ), chosen to reflect plausible dispersion values

Inference was conducted using:

- 4 chains per model  
- 1000–1500 tuning steps  
- ≥2000 posterior samples per chain

I assessed convergence using R-hat (target < 1.01) and effective sample size (ESS > 1000). Full trace plots and diagnostics are included in the Supplement.

@@

I compared model performance using **Pareto Smoothed Importance Sampling Leave-One-Out Cross-Validation (PSIS-LOO-CV)**, implemented via the `arviz.loo()` function. This Bayesian metric estimates the expected log predictive density (ELPD), a principled measure of out-of-sample predictive accuracy.

>>>>

## Materials and Methods

@@

### Data Description and Preparation Overview

@@

### Bayesian Modeling Framework

@@

*Model 2* introduces partial pooling across four groups defined by the dominant MBR numerator band; Rrs411, Rrs443, Rrs489, Rrs510 (*cf*. @fig-eda-mbr). Each group receives its own slope and intercept with shared hyperpriors. On the one hand each group has its own sub-model that can capture details specific to it, while the shared hyperpriors allow for the sharing of information between groups when fitting the model, thus maximizing the use of information contained in the available data.  Mathematical description of the model can be found in the Supplementary Material. @fig-model2-struct illustrates the model structure.

@@

Model 2 - Multi-level hierarchical level with partial pooling across $MBR$ numerator groups (Rrs411, Rrs443, Rrs489, Rrs510). Model parameter prior mean and variance have Gaussian and Exponential hyperpriors. This structure allows the sharing of information across groups at model fitting time.

@@

The data plotted in @fig-eda suggests heteroscedasticity - variance that is not constant, but rather changes depending on the data. Models 3 and 4 investigate the role of group-wise constant heteroscedasticity and global linear heteroscedasticity, respectively. Details of these models can be found in the Supplementary Material section. To investigate this further, Model 5 looks at group-wise linear heteroscedasticity. Thus, for a given MBR group $j$ $σ_j$ is modeled as a log-linear relationship such that $$log(σ) = α_j + β_j × log(MBR)$ and then transformed via $exp(log(σ_j))$ before being passed on to the likelihood. Similar to the model mean,  $μ_j$, $σ_j$'s slope and intercept terms are partially pooled and so have common hyperpriors. Mathematical descriiption of the model can be found in the Supplementary Material section; @fig-model5-struct.  illustrates the model structure.

@@

::: {#fig-model6-struct fig-pos="H"}

@@

### Prior Specifications and Model Fitting

*Priors* - I used weakly informative priors across all models to balance flexibility and regularization.  For regression parameters  I used Gaussian priors, which in the case of single-level models like Model 1, were given a $0$ mean and variance $1$, to encourage parameter sparsity. Hyperpriors of prior means were set to Gaussans of mean 0 and variance 1 to encourage sparsity. Hyperpriors of prior variances set to  Exponential distribution with rate set to $$1$, which has significant density near 0, encourages similarity across groups unless the data mandates otherwise. This hyperprior also encourages uncertainty shrinkage across groups if possible. The likelihood dispersion $σ$, when constant was set, to a Gamma distribution of shape and rate both set to $2$. Prior soundness was assessed using *Prior Predictive Checks*. This procedure takes advantage of the fact taht Bayesian models are generative to insure that values of ($log(Chl)$) predicted by the model are reasonable even before model fitting. This was plotted as Cumulative Distribution Function (CDF) of $log(Chl)$. See *Results* section for a Prior Predictive Checks plot example; plots for all models can be found in the Supplementary Material section.

*Model Fitting* - The model posterior was approximated using the No U-Turn Sampler (NUTS), an efficient and mature variant of Hamiltonian Monte Carlo (HMC), which uses particle physics to sample the posterior. Sampler settings were as follows. Four independent sampling runs referred to as *chains* were used to assess sampler convergence. Prior to sampling, 1000 tuning steps were used for each chain to maximize sampler efficiency. The 2000 samples were recorded for each chain totalling 8000 samples. I assessed sampler convergence using the Gelman-Rubin statistic, $\hat{R}$ and effective sample size($ESS$). Convergence is acceptable for $\hat{R}$ < 1.01 and $ESS > 1000$. Full trace plots and diagnostics are included in the Supplementary Material section. Goodness of fit was assessed using *Posterior plots*, which show  *Posterior Predictive Checks, which show the expected

@@

I compared model performance using **Pareto Smoothed Importance Sampling Leave-One-Out Cross-Validation (PSIS-LOO-CV)** [@vehtari2017]. This Bayesian metric estimates the expected log predictive density (ELPD), a principled measure of out-of-sample predictive accuracy, that uses the entire posterior predictive distribution
